Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 Mw=7.9 Wenchuan earthquake

### 2014-09-17 paper.tex
Added to discussion section.

\section{Introduction}
The Longmen Shan range is located on the eastern margin of the Tibetan and rises 4000m over 30km.  The 2008 $\mathrm{M_w}$ 7.9 Wenchuan earthquake ruptured 250km along strike on the Beichuan and Pengguan faults, resulting in approximately 70000 fatalities. Prior geodetic estimates suggest almost zero horizontal shortening \citep{king97, chen00, shen05, Meade07c, Loveless2011}, in contrast with the dominant thrust slip-sense during the Wenchuan rupture.  Other authors reconcile the lack of shortening with the steep topography by appealing a lower crustal inflation tectonic model for the eastern Tibetan Plateau \citep{royden97, bird91, Burchfiel2008a}.  However, the magnitude and location of the Wenchuan earthquake suggest a more typical fold-and-thrust belt.  Using seismic reflection data, \citet{hubbard09} find a well-developed fold-and-thrust fault system geometry with long-term shortenings up to 100\%.  A growing body of literature supports this interpretation. \citet{Fielding2012a} interpret satellite gravity-field data to infer that mass of the Longmen Shan is supported by a flexed Sichuan Basin.  This fold-and-thrust structure may extend deep into the plateau as suggested by seismic tomography, reflection, magnetotelluric data and studies of the Longriba fault zone 200km northwest of the Longmen Shan \citep{Xu2008, Zhang2009, Zhao2012, Ren2013, Guo2013}. (LOTS TO EXTEND HERE, SPACE?)

@@

\section{Implications for the Longmen Shan earthquake cycle}
The striking feature of Figure \ref{fig:big_stack}a is the lack of a range-front geodetic signature of shortening.

Model without topography do not have a local minimum northwest of the Beichuan, indicating the relevance of topography in crustal deformation.

This far-field deformation is consistent with vertical leveling data that suggests a vertical velocity profile that increases from the Longmen Shan range front until peaking 50km southeast of the Longriba fault zone \citep{Hao2014}. The Longriba fault zone has the wrong slip-sense (up to the south-east, \citep{Ren2013}) to accomodate such vertical deformation. No other structure is nearby.

Figure \ref{fig:hazard} shows a summary of recurrence intervals as a function of total event slip. Our shortening estimate suggests a recurrence interval for Wenchuan-like ruptures of approximately 600 years. Recurrence for smaller events is much more frequent. This contrasts with the average recurrence of ~2000 years determined by \citep{Ran2010} for the Beichuan fault. Our estimate would predict $M_w = $9.0 ruptures on that recurrence, seemingly unlikely given the total slip and fault length required. However, our estimate is an system-wide recurrence while that paleoseismic estimate is limited by only studying one of multiple imbricated thrusts.

There are reasons to believe that the seismic hazard estimate presented here is a low estimate.  First, coseismic and long term slip estimates include (CITE VARIOUS COSEISMIC + DENSMORE) a significant component of strike-slip motion. But, we only model horizontal shortening.  Second, our two-dimensional model assumes plane strain conditions.  The approximation will result in under-estimates of the shortening required to produce far-field velocities.  This is because the assumption is equivalent to assuming the fault geometry extends infinitely far along strike, resulting in a fault surface much larger than in reality. Third, the modeled thrust is the steeply dipping Beichuan fault. The foreland Range Front thrust or the shallow detachment beneath Chengdu dip more shallowly \citep{Hubbard2010}, producing a greater total moment deficit.  Finally, the shortening rate estimate may be low due to being estimated late in the earthquake cycle \citep{savage00}.  It is also important that the recurrence interval calculation assumed that the entire range-front was ruptured by the Wenchuan earthquake whereas the southern portion of the Longmen Shan did not rupture during the earthquake and is thought to similarly active to the portions that did rupture \citep{Li2010}.

@@

Previous near-zero shortening estimates in the Longmen Shan are at odds with the large-moment of the Wenchuan earthquake, very steep topography, structural interpretations indicate fold-and-thrust geometry, fast erosion rates. We clarify this conflict by including earthquake cycle effects and a 20km deep detachment. Our revised interseismic shortening rate is 5.7 $\pm$ 1.5 mm/yr. This interpretation unifies models of coseismic (Wenchuan), postseismic and interseismic  based on geodetic data. Our model suggest that Wenchuan-like events have much shorter recurrence intervals.

>>>>

\section{Introduction}
The Longmen Shan range is located on the eastern margin of the Tibetan and rises 4000m over 30km.  The 2008 $\mathrm{M_w}$ 7.9 Wenchuan earthquake ruptured 250km along strike on the Beichuan and Pengguan faults, resulting in approximately 70000 fatalities. Prior geodetic estimates suggest almost zero horizontal shortening \citep{king97, chen00, shen05, Meade07c, Loveless2011}, in contrast with the dominant thrust slip-sense during the Wenchuan rupture.  Other authors reconcile the lack of shortening with the steep topography by appealing a lower crustal inflation tectonic model for the eastern Tibetan Plateau \citep{royden97, bird91, Burchfiel2008a}.  However, the magnitude and location of the Wenchuan earthquake suggest a rapidly shortening fold-and-thrust belt. From seismic reflection data, \citet{hubbard09} find a well-developed fold-and-thrust fault system geometry with long-term shortenings up to 100\%. \citet{Fielding2012a} interpret satellite gravity-field data to infer that mass of the Longmen Shan is supported by a flexed Sichuan Basin. Seismic tomography, seismic reflection and magnetotelluric data have been used to suggest that this fold-and-thrust structure extends deep into the plateau as a deep detachment \citep{Zhang2009, Zhao2012, Guo2013}. Erosion rates are The Longriba fault zone, 200km northwest of the Longmen Shan, is suggested to also sole into this detachment \citep{Xu2008, Ren2013}.

@@

\section{Implications for the Longmen Shan earthquake cycle}
The striking feature of Figure \ref{fig:big_stack}a is the lack of a range-front geodetic signature despite the significant shortening rate. Locking on the deep detachment obscures the signal of deformation on the range-front faults. According to this model, interseismically, the eastern Tibetan Plateau moves cohesively with the Sichuan Basin, whereas coseismic slip is concentrated at the boundary between these two tectonic blocks. Traditionally, the velocity gradient we attribute to a deep detachment has been attributed to broad viscous strain across the eastern Tibetan Plateau (CITE).

Models without surface topography do not have a local minimum northwest of the Beichuan, indicating the relevance of topography in crustal deformation. This effect may be more pronounced with a shallower fault dip like in many subduction zones.

Our model of far-field deformation is consistent with a similar detachment model of leveling and GPS-based uplift rates \citep{Hao2014}. The vertical velocities increase from the Longmen Shan range front until peaking 50km southeast of the Longriba fault zone at approximately 3.5 mm/yr of uplift. The Longriba fault zone has the wrong slip-sense (up to the south-east, \citep{Ren2013}) to accomodate such vertical deformation. No other identified structures are nearby.

\citet{kirby07} study an eastward decrease in interseismic velocities across the Kunlun fault. They conclude that the decline in slip-rate must be attributed to internal deformation. However, the same feature could be explained with slip being accomodated by a wide eastern Tibetan fold-and-thrust system, a suggestion supported by the continuation of shortening north from the Longmen Shan onto the Huya \citep{kirby00} and Min Jiang faults \citep{Chen1994}.

We have not addressed when the accumulated slip-deficit on the deep detachment is released. Postseismic slip incorporated in the model of \cite{Qi2011} shows up to 5 meters of slip on a deep detachment. Our data analysis depends on the detachment being locked during the period of data collection (WHEN?). Aseismic slip may nonetheless be the predominant mode of strain release given the short observed portion of the earthquake cycle. Seismicity is approximately limited to upper 20km of crust in the range-front \citep{Li2010}. On the other hand, the surface area and shortening rate of the detachment could sustain very large earthquakes if slip is released coseismically.

Using steady-state critical taper wedge theory, we estimate a basal coefficient of friction between 0.02 and 0.30 for the detachment. The mean parameter estimate of 0.05 is similar to estimates of friction on the low-angle detachment system beneath the Longmen Shan foreland basin \citep{Hubbard2010}.

Figure \ref{fig:hazard} shows a summary of recurrence intervals as a function of total event slip. Our shortening estimate suggests a recurrence interval for Wenchuan-like ruptures of approximately 600 years. Recurrence for smaller events is much more frequent. This contrasts with the average recurrence of ~2000 years determined by \citep{Ran2010} for the Beichuan fault. Our estimate would predict $M_w = $9.0 ruptures for that recurrence, seemingly unlikely given the total slip and fault length required. However, our estimate is an system-wide recurrence while paleoseismic estimates have been limited by only studying one of multiple imbricated thrusts.

There are reasons to believe that the seismic hazard estimate presented here is a low estimate.  First, coseismic and long term slip estimates include a significant component of strike-slip motion\citep{Shen2009, Qi2011, Densmore2007} . But, we only model horizontal shortening.  Second, our two-dimensional model assumes plane strain conditions.  The approximation will result in under-estimates of the shortening required to produce far-field velocities.  This is because the assumption is equivalent to assuming the fault geometry extends infinitely far along strike, resulting in a fault surface much larger than in reality. Third, the modeled thrust is the steeply dipping Beichuan fault. The foreland Range Front thrust or the shallow detachment beneath Chengdu dip more shallowly \citep{Hubbard2010}, producing a greater total moment deficit.  Finally, the shortening rate estimate may be low due to being estimated late in the earthquake cycle \citep{savage00}.  It is also important that the recurrence interval calculation assumed that the entire range-front was ruptured by the Wenchuan earthquake whereas the southern portion of the Longmen Shan did not rupture during the earthquake and is thought to similarly active to the portions that did rupture \citep{Li2010}.

@@

Previous near-zero shortening estimates in the Longmen Shan are at odds with the large-moment of the Wenchuan earthquake, very steep topography, structural interpretations indicate fold-and-thrust geometry, fast erosion rates. We clarify this conflict by including earthquake cycle effects and a 20km deep detachment. Our revised interseismic shortening rate is 5.7 $\pm$ 1.5 mm/yr. The model shows that interseismic strain accumulation may be spatially disjoint with coseismic strain release, emphasizing that accurate fault geometries are critical in the analysis of geodetic data. This interpretation unifies geodetic models of coseismic (Wenchuan), postseismic and interseismic crustal deformation in the Longmen Shan. Finally, our model suggest that Wenchuan-like events have much shorter recurrence intervals.
